Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-20 article/main.tex
dessin + début section 5

\documentclass[preprint]{iacrtrans}

@@

Each time the PRNG is clocked, 64 output bits are extracted from the internal
state using a non-linear function that makes use of data-dependent rotations, in
a way reminiscent of the \textsf{RC5} block cipher~\cite{Rivest94}. The six most
significant bits of the internal state encode a number $0 \leq r < 64$. The two
64-bit halves of the internal state are XORed together, and this 64-bit result
is rotated right by $r$ positions. The successive 64-bit outputs of the
generator are $X_0, X_1, \dots$ where:
\[
  X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
\]

@@

\section{C connu}

\subsection{Réduction}
Pour deviner l'état interne du générateur, une méthode consiste à procéder par brute-force. On devine les 64 bits supérieurs de $S_0$, on en déduit la rotation $r_0$ utilisée, on peut retrouver  les 64 bits inférieurs:

\begin{equation}
   S_0[0:64] = S_0[64:128] \oplus \texttt{ROT}(X_0 , 64 - r_0) \\
\end{equation}

Cette méthode demanderait de l'ordre de $2^{64}$ calculs, ce qui n'est pas acceptable. Nous pouvons cependant utiliser cette équation pour déduire des informations sur l'état interne de S. En effet,en devinant par brute-force les 6 bits du haut de S, et ses $\ell$ bits du bas, on récupère :

\todo{schéma "relevant bits of the outputs"}

On réduit $S$ à une série géométrique plus simple. Notons $S'$ l'état interne auquel on retire sa composante en $C$, et en $S_0[0:\ell]$ (note de bas de page : le fait de retirer $S_0[0:\ell]$ n'est pas nécéssaire, mais nous permet de limiter la taille des instances utilisées à 64 bits, très utile pour la programmation en C). Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$. On se réduit donc au cas facile étudié dans la partie précédente.
\begin{equation}
genMatrix =
\begin{pmatrix} 
1 & A & A ^2 & \dots & A ^{n- 1}\\
0 & M & 0 & \dots & 0\\
0 & 0 & M & \dots & 0\\
\vdots & & & \ddots & \vdots\\
0 & 0 & \dots & 0 & M
\end{pmatrix}
\end{equation}

\subsection{Algorithme}
On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau $\mathcal{L}(A,2^64,n)$ par l'algorithme LLL.
Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
\begin{enumerate}
\item On récupère le $T$ correspondant à nos suppositions. On calcule $T'$. ($T'_i = (S_i[64 - 6 : 64 + \ell] - (A^i*S_0[0 : \ell] + C * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[64 - 6 : 64 + \ell] * 2^{64 - 6 - \ell}$)

>>>>

\documentclass[preprint,svgnames]{iacrtrans}

@@

Each time the PRNG is clocked, 64 output bits are extracted from the internal
state using a non-linear function that makes use of data-dependent rotations, in
a way reminiscent of the \textsf{RC5} block cipher~\cite{Rivest94}. The six most
significant bits of the internal state encode a number $0 \leq r < 64$. The two
64-bit halves of the internal state are XORed together, and this 64-bit result
is rotated right by $r$ positions. The successive 64-bit outputs of the
generator are $X_0, X_1, \dots$ where:
\begin{equation}\label{eq:output}
  X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
\end{equation}

@@

\section{Warm-up: A Prediction Algorithm for \textsf{PCG64} With Known Increment}

We first consider the easier case where the ``increment'' (the $c$ term in the
definition of the underlying linear congruential generator) is known --- recall
that a default value is specified in case the user of the pseudo-random
generator does not want to provide one.

In this case, reconstructing the 128-bit internal state $S_i$ of the generator
obis sufficient to produce the pseudo-random flow with 100\% accuracy (the
generator can also be clocked backwards if necessary, so that the seed can be
easily reconstructed). We therefore focus on $S_0$, the first internal state
after seeding. A very simple strategy could be the following:
\begin{enumerate}
\item Guess the 64 upper bits of $S_0$.
\item Compute the missing 64 lower bits using~\eqref{eq:output}, with:
\[
   S_0[0:64] = S_0[64:128] \oplus (X_0  \lll S[122:128]).
\]
\item If $X_1$ is computed correctly, output $S_0$.
\end{enumerate}

This requires $2^{64}$ iterations of a loop that does a dozen arithmetic
operations. This is the baseline that we improve upon. An improved
``guess-and-determine'' prediction algorithm is possible, which essentially
amounts to expose a truncated version of the underlying linear congruential
generator, and attack it using the well-known tools. This is possible by
combining the following ingredients:
\begin{itemize}
\item The underlying linear congruential generator uses a power-of-two modulus,
  therefore the $\ell$ low-order bits of $S_{i+1}$ are entirely determined by
  the $\ell$ low-order bits of $S_i$. More precisely, we have:
  \begin{equation}\label{eq:lcg}
    S_{i+1} = aS_i + c \bmod 2^\ell, \qquad \text{for all } 0 \leq \ell \leq 128
  \end{equation}
  Therefore, guessing low-order bits of $S_0$ yields a ``long-term advantage''
  that holds for all subsequent states.

\item Guessing a 6-bit rotation gives access to the XOR of the two halves of the
  internal state. If a part of the state is known, then this transfers existing
  knowledge to the other half.
\end{itemize}

In figure~\ref{fig:Cknown}, we see that guessing $S_0[0:\ell]$ and a few 6-bit
rotations give access to $S_i[58:64+\ell]$ for the corresponding consecutive
states. Therefore, we are facing a truncated linear congruential generator on
$64+\ell$ bits, where we only know $6+\ell$ bits of output, for a few
consecutive outputs. This is sufficient to reconstruct entirely the successive
states of this truncated linear congruential generator. This reveals
$S_0[0:64+\ell]$, and using~\eqref{eq:output} the entire $S_0$ can be
reconstructed. The precise details follow.

\begin{figure}
\begin{center}
\begin{tikzpicture}[yscale=0.66]
  \path[red, use as bounding box] (-0.75, -9.5) rectangle (12.8, 2.25);

  % S_0
  \begin{scope}
    % remplissage
    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
    \fill[fill=magenta] (11, 0) rectangle  +(1.8, 1);

    % déco autour
    \node at (-0.5, 0.5) {$S_0$};
    \draw[<->] (0, 1.5) -- node[above] {64 bits} +(6.375, 0);
    \draw[<->] (6.425, 1.5) -- node[above] {64 bits} +(6.375, 0);
  \end{scope}

  % T_0
  \begin{scope}[xshift=4.6cm, yshift=-1.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);

    % déco
    \node at (-0.5, 0.5) {$T_0$};
  \end{scope}

  %%%%%%%%%

  % S_1
  \begin{scope}[yshift=-3cm]
    % remplissage
    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
    \fill[fill=cyan] (11, 0) rectangle +(1.8, 1);

    % déco autour
    \node at (-0.5, 0.5) {$S_1$};

    % flèches S_i --> T_i
    \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
    \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
  \end{scope}

  % T_1
  \begin{scope}[xshift=4.6cm, yshift=-4.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);

    % déco
    \node at (-0.5, 0.5) {$T_1$};
  \end{scope}

  %%%%%%%%%%%%%

  % S_2
  \begin{scope}[yshift=-6cm]
    % remplissage
    \fill[fill=Yellow] (4.6, 0) rectangle +(2.4, 1);
    \fill[fill=magenta] (0, 0) rectangle +(0.6, 1);
    \fill[fill=cyan] (11, 0) rectangle +(1.8, 1);

    % déco autour
    \node at (-0.5, 0.5) {$S_2$};

    % flèches S_i --> T_i
    \draw[ultra thick,->] (5.5, 0.5) -- +(0, -1.5);
    \draw[ultra thick,->] (6.7, 0.5) -- +(0, -1.5);
  \end{scope}

  % T_2
  \begin{scope}[xshift=4.6cm, yshift=-7.5cm]    
    \draw[dashed]  (0, 0) rectangle +(6.4, 1);
    \draw[thick]  (0, 0) rectangle +(2.4, 1);
    \draw[]  (1.8, 0) rectangle +(0, 1);
    \path (2.4, 0) rectangle node {$00000000 \cdots 00000000$} (6.4, 1);

    % déco
    \node at (-0.5, 0.5) {$T_2$};
    \draw[<->] (0, -0.5) -- node[below] {$\ell$ bits} +(1.775, 0);
    \draw[<->] (1.8, -0.5) -- node[below] {6} +(0.6, 0);
    \draw[<->] (0, -1.25) -- node[below] {64 bits} +(6.4, 0);
    \draw[<->] (6.4, -1.25) -- node[below] {$\ell$ bits} +(1.775, 0);
  \end{scope}
\end{tikzpicture}

\end{center}
\caption{A guess-and-determine algorithm to reconstruct the first internal state
  $S_0$. Magenta bits are guessed; cyan bits are obtained using the linear
  congruence relation~\eqref{eq:lcg}; yellow bits are obtained
  from the output and the guessed rotations using~\eqref{eq:output}.}
\label{fig:Cknown}
\end{figure}

Set $S_i' = S_i - (S_0[0:\ell] + c) \sum_{j=0}^i a^i$. It is easy to check that
$S'_{i+1} = a S'_i$, and therefore the successive $S'_i$'s follow a geometric
progression\footnote{note de bas de page : le fait de retirer $S_0[0:\ell]$
  n'est pas nécéssaire, mais nous permet de limiter la taille des instances
  utilisées à 64 bits, très utile pour la programmation en C)}.

Nous utiliserons le réseau engendré par les $S'[\ell : 64 + \ell]$, nommé $\mathcal{L}(A,2^64,n)$. On se réduit donc au cas facile étudié dans la partie précédente.
\begin{equation}
genMatrix =
\begin{pmatrix} 
1 & A & A ^2 & \dots & A ^{n- 1}\\
0 & M & 0 & \dots & 0\\
0 & 0 & M & \dots & 0\\
\vdots & & & \ddots & \vdots\\
0 & 0 & \dots & 0 & M
\end{pmatrix}
\end{equation}

On calcule préalablement les matrices $G$ et $G^{-1}$, avec $G$ la matrice génératrice réduite du réseau $\mathcal{L}(A,2^64,n)$ par l'algorithme LLL.
Pour tous les $r_i$ et les $S_0[0 : \ell]$ possibles:\\
\begin{enumerate}
\item On récupère le $T$ correspondant à nos suppositions. On calcule $T'$. ($T'_i = (S_i[64 - 6 : 64 + \ell] - (A^i*S_0[0 : \ell] + C * \sum_{j = 0}^i A^j)[64 - 6 : 64 + \ell])[64 - 6 : 64 + \ell] * 2^{64 - 6 - \ell}$)
